---
title: "Project3: Data Science Skills"
author: "Madina Kudanova, Muhammad Khan, Leon Lin, Xiaofei Mei"
format: html
editor: visual
---

```{r setup, message=FALSE, include=FALSE}
# Load packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(DBI)
library(RPostgres)
library(dplyr)
library(tidyr)
library(stringr)
library(janitor)
```

## Introduction

The project Most Valued Data Science Skills aims to investigate which technical skills are most valued in the data science job market. As the demand for data science professionals has grown significantly over the last decade, it is important to understand which tools, programming languages, and analytical skills are most frequently required by employers.

To explore this question, the project will analyze datasets containing job postings for data science related roles. Job postings provide insight into employer expectations because they typically list the technical skills required for each position. In this project, the frequency of skills listed in job postings will be used as a proxy for measuring the demand for specific data science skills.

## Approach

### Team members

-   Madina Kudanova

-   Muhammad Khan

-   Leon Lin

-   Xiaofei Mei

### Communication

Our team will communicate primarily through WhatsApp/Slack group chats and school email to coordinate meetings, discuss progress, and share quick updates related to the project.

### Code and File Sharing

A shared GitHub repository <https://github.com/MKudanova/Project3_DataScienceSkills/tree/main> will be used to store project code, scripts, and datasets.

### Project Documentation

A shared Google Docs document will be used to collaboratively write project notes and draft written sections of the project before final versions are uploaded to the GitHub repository.

### Data Sources

We found the data from Kaggle. The data was scrapped from Glassdoor with information such as job title, job description, company information, industry. Skills required mostly can be listed through th job description column.

### Data Loading and Database Design

After obtaining the dataset, it will be downloaded in CSV format and loaded into a relational database. The database will follow a normalized structure to organize the data efficiently. Because one job posting can require multiple skills and a single skill may appear in multiple job postings, the database will use a many-to-many relationship between jobs and skills. The initial design might include three tables: Jobs (job_id, job_title, company, location), Skills (skill_id, skill_name), and Job_Skills (job_id, skill_id). The data will then be accessed from R for cleaning, tidying, and exploratory analysis. The database structure may be adjusted. The relational database will be implemented using PostgreSQL, managed through pgAdmin. SQL scripts used to create the database schema, tables, and relationships will be included in the project’s GitHub repository to ensure reproducibility and transparency of the data loading process.

### ER Diagram

To document the structure of the database, the project will include an Entity–Relationship (ER) diagram illustrating the relationships between the tables in the normalized database. It will represent the relationships between the jobs and skills tables. The diagram will most likely be created using a visual diagramming tool such as draw.io (diagrams.net) or another diagram tool. The ER diagram may be refined after the dataset is examined and the final database schema is implemented.

### Exploratory Analysis and Visualization

After the data are cleaned and transformed in R, exploratory data analysis will be conducted to examine the frequency of different data science skills across job postings. Visualizations such as bar charts, frequency plots, and comparative graphs will be created to illustrate which programming languages, tools, and technologies appear most frequently in job listings. These visualizations will help communicate the relative demand for different data science skills in an intuitive and interpretative way.

### Load data tables from local database

For safety concern, we will first set up environmental variable to store password and use it to establish connection.

```{r load_data, message=FALSE}
con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "project3_datascience_skills",
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = Sys.getenv("PGPASSWORD")   
)

```

### Data Preview and Validation

In database, there is a view been set up through the attached SQL code. It will be the main dataset we use for the following analysis part. Here is a preview:

```{r}

job_skill_details <- dbGetQuery(con, "SELECT * FROM project3.vw_job_skill_details")

head(job_skill_details)

names(job_skill_details)

```

### Skill Demand Analysis

```{r}
# Plot the frequency of each skill across all job postings
ggplot(skill_counts_df, aes(x = reorder(skill_name, n), y = n)) +
  geom_col(fill = "orange") +
  coord_flip() +
  theme_minimal() +
  labs(title = "                                  Most In-Demand Skills  ", x = " Skills ", y = "
       Skill Popularity")
```

### Skill Demand Across Salary Levels

```{r}
#salary brackets based on data distribution
salary_quantiles <- quantile(jobs$avg_salary, probs = c(0, 0.2, 0.4, 0.6, 0.8, 1), na.rm = TRUE)

# Create salary brackets 
salary_analysis <- job_skill_details |>
  filter(!is.na(avg_salary)) |>
  mutate(
    salary_bracket = case_when(
      avg_salary <= salary_quantiles[2] ~ sprintf("≤ $%.0f", salary_quantiles[2]),
      avg_salary <= salary_quantiles[3] ~ sprintf("$%.0f - $%.0f", salary_quantiles[2], salary_quantiles[3]),
      avg_salary <= salary_quantiles[4] ~ sprintf("$%.0f - $%.0f", salary_quantiles[3], salary_quantiles[4]),
      avg_salary <= salary_quantiles[5] ~ sprintf("$%.0f - $%.0f", salary_quantiles[4], salary_quantiles[5]),
      TRUE ~ sprintf("> $%.0f", salary_quantiles[5])
    )
  )

# Get top 5 salary brackets
top_brackets <- salary_analysis |>
  count(salary_bracket, sort = TRUE) |>
  slice_head(n = 5) |>
  pull(salary_bracket)

# Find top 4 skills by salary bracket
top_skills_by_salary <- salary_analysis |>
  filter(salary_bracket %in% top_brackets) |>
  count(salary_bracket, skill_name, sort = TRUE) |>
  group_by(salary_bracket) |>
  slice_max(n, n = 4, with_ties = FALSE) |>
  ungroup() |>
  arrange(salary_bracket, desc(n))


#Heatmap - Top Skills Across Salary Brackets
ggplot(top_skills_by_salary, 
       aes(x = salary_bracket, y = reorder(skill_name, n), fill = n)) +
  geom_tile(color = "white", size = 1) +
  scale_fill_gradient(low = "#FEE5D9", high = "#A50F15", 
                      name = "Frequency",
                      labels = scales::comma) +
  geom_text(aes(label = n), color = "black", size = 3.5, fontface = "bold") +
  labs(
    title = "Top Skills Across Different Salary Brackets",
    subtitle = paste("Based on", nrow(salary_analysis), "job postings with salary data"),
    x = "Salary Bracket",
    y = "Skill"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.position = "right"
  )

# PLOT 2:Bar ChartMost In-Demand Skills by Salary Bracket
ggplot(top_skills_by_salary, 
       aes(x = reorder(skill_name, n), y = n, fill = salary_bracket)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2, size = 3) +
  coord_flip() +
  facet_wrap(~salary_bracket, scales = "free_y", ncol = 2) +
  labs(
    title = "Most In-Demand Skills by Salary Bracket",
    subtitle = "Top 4 skills within each salary range",
    x = "Skill",
    y = "Number of Job Postings"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    strip.text = element_text(face = "bold", size = 10),
    axis.text = element_text(size = 9),
  )


```

### 

## Skill Demand by Sector

```{r}
top_sectors <- jobs |>
  filter(sector != "-1", !is.na(sector), sector != "") |>
  
  # Count number of job postings per sector and sort
  count(sector, sort = TRUE) |>
  
  # Keep the top 4 
  slice_head(n = 4)

job_skill_salary <- job_skills |>
  left_join(jobs, by = "job_id") |>
  left_join(skills, by = "skill_id")


job_skill_sector <- job_skill_salary |>
  filter(sector %in% top_sectors$sector)

top_skills_by_sector <- job_skill_sector |>
  count(sector, skill_name, sort = TRUE) |>
  group_by(sector) |>
  slice_max(n, n = 4, with_ties = FALSE) |>
  
  # Remove grouping for clean output or further analysis
  ungroup()

print(top_skills_by_sector)
sector_ratio <- jobs |>
  filter(sector != "-1", !is.na(sector), sector != "") |>  
  count(sector) |>                        
  mutate(percentage = n / sum(n) * 100)       

# Round percentages 
sector_ratio_clean <- sector_ratio |>
  mutate(percentage = round(percentage, 0)) |>
  select(sector, percentage)

# Sort sectors from highest to lowest
sector_ratio_sorted <- sector_ratio_clean %>%
  arrange(desc(percentage))

print(sector_ratio_sorted)

#plot
ggplot(top_skills_by_sector,
       aes(x = reorder(skill_name, n), y = n)) +
  
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), hjust = -0.1, size = 3) +
  
  coord_flip() +
  facet_wrap(~ sector, scales = "free_y") +
  
  labs(
    title = "Top Skills Across Leading Job Sectors",
    x = "Skill",
    y = "Number of Job Postings"
  ) +
  theme_minimal() +
  
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    strip.text = element_text(face = "bold", size = 10),
    axis.text.y = element_text(size = 9)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

#### 

## Conclusion

Based on the Glassdoor dataset analyzed, which was transformed into a relational database using SQL, core data science skills, particularly Python, machine learning and statistical analysis consistently emerge as the most in-demand. These skills appear across all salary ranges and industry sectors represented in the dataset, indicating that their value is stable and widely recognized regardless of role or domain. Looking from sector distribution, excel get to mentioned in the top 4 in business services sector but didn't make to the top 4 in sector information technology and biotech pharmaceuticals.

Overall, the findings suggest that a strong foundation in key programming and analytical skills forms a reliable “value pack” for data scientists.
